Time Series Modeling of Varying Complexity Captures Subdiffusive Solute Dynamics and Predicts Long Timescale Behavior in Nanoscale Pores.

### 2019-12-06 Ben_Manuscripts/stochastic_transport/Draft.tex
New flux curves

  We determined the rate at which solutes cross macroscopic-length pores based on the
  Hill relation:~\cite{hill_free_1989}
  \begin{equation}
  J = \frac{1}{MFPT}
  \end{equation}
  where $J$ is solute flux and MFPT refers to the mean first passage time. In the 
  context of our work, the MFPT describes the average length of time it takes a particle
  to move from the pore entrance to the pore exit.

@@

  First, consider an initial pulse spreading out over time with a fixed mean. We can
  solve for the time-dependent probability density of particle positions by solving 
  the one dimensional diffusion equation:
  \begin{equation}
  \frac{\partial p}{\partial t} = D \frac{\partial^2 p}{\partial z^2}
  \end{equation}
  The appropriate initial and boundary conditions are:
  \begin{equation}
  BC1: t > 0, z = \infty, p = 0
  \end{equation}
  \begin{equation}
  BC2: t > 0, z = 0, \frac{\partial p}{\partial z} = 0
  \end{equation}
  \begin{equation}
  IC: t = 0, c = M\delta(z)
  \end{equation}
  where $M$ is the total number of particles in the system. It has been shown 
  elsewhere that the solution to this equation is:
  \begin{equation}
  p(z, t) = \frac{1}{\sqrt{4 \pi D t}}exp\bigg(\frac{-z^2}{4Dt}\bigg)
  \end{equation}
  We can make the substitution $z = z - vt$ in order to linearly shift the mean as a
  function of time:
  \begin{equation}
  p(z, t) = \frac{1}{\sqrt{4 \pi D t}}exp\bigg(\frac{-(z - vt)^2}{4Dt}\bigg)
  \end{equation}
  One can track the fraction of particles, $F$, that have crossed the pore 
  boundary by integrating:
  \begin{equation}
  F = \int_L^\infty p~dz = erfc\bigg(\frac{L - vt}{2\sqrt{D t}}\bigg)
  \end{equation}
  where $L$ is the pore length. This represents the cumulative first passage 
  time distribution so we take its derivative in order to arrive at the first
  passage time distribution:
  \begin{equation}
  P = -\frac{1}{\sqrt{\pi}}e^{-(L - vt)^2 / (4Dt)}\bigg(-\frac{D(L - vt)}{4(Dt)^{3/2}} - \frac{v}{2\sqrt{Dt}}\bigg)
  \label{eqn:passage_times}
  \end{equation}

@@

  %BJC4: another way to visualize data
  %BJC4: I think these are kind of hard to read. Can't really see any trends. Open to other
  %ideas.
  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\linewidth]{1mode_radar.pdf}
  \caption{}\label{fig:1mode_radar}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\linewidth]{2mode_radar.pdf}
  \caption{}\label{fig:2mode_radar}
  \end{subfigure}
  \caption{Another way to visualize Tables~\ref{table:sfbm_params} and~\ref{table:sfbm_params_2mode}}\label{fig:radars}
  \end{figure}

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\linewidth]{1mode_radar_nolines.pdf}
  \caption{}\label{fig:1mode_radar}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\linewidth]{2mode_radar_nolines.png}   % for some reason it has trouble saving this as a PDF
  \caption{}\label{fig:2mode_radar}
  \end{subfigure}
  \caption{Same as Figure~\ref{fig:radars} but with no lines between points}\label{fig:radars_nolines}
  \end{figure}

@@

  \subsection{Solute Mean First Passage Times}\label{section:mfpt}

  We used the 1-mode sFBMcut (see Table~\ref{table:anomalous_models}) anomalous 
  diffusion model in order to demonstrate how one can use its realizations in
  order to calculate the MFPT (section~\ref{method:mfpt}) of solutes given model 
  parameters extracted from MD simulations. Any approach which allows one to 
  generate realizations of solute trajectories can be used in place of the AD model.

  In Figure~\ref{fig:mfpt_curves_regular} we show that the MFPT grows super-linearly
  with pore length. We fit a curve of the form $cL^{\alpha}$ to the curves in order
  to estimate the growth exponent, $\alpha$. The growth exponent is inversely proportional
  to the solute MSDs which is consistent with longer passage times for solutes that
  progress through the pores more slowly.

  The growth exponent is inversely related to the hurst parameter, $H$. In 
  Figure~\ref{fig:mfpt_curves_brownian}, we removed the correlation structure (set
  $H$=0.5) from AD trajectory realizations so that all hops were uncorrelated.
  Regardless of the parameters of the dwell time distribution, the growth 
  exponent is always about 2, the same value that we observe for Brownian 
  motion (see Figure~\ref{S-fig:mfpt_curve_brownian} of the Supporting 
  Information).

  % BJC5: I'm not sure the best way to calculate statistics on this. Since I don't know
  % the density function of the mean first passage time distribution, I use bins to 
  % to locate the peaks. So the error bars are dependent on the bin width. 
  \begin{figure}
  \centering
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{mfpt_curves.pdf}
  \caption{Correlated Hops}\label{fig:mfpt_curves_regular}
  \end{subfigure}
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{mfpt_curves_brownian.pdf}
  \caption{Uncorrelated Hops}\label{fig:mfpt_curves_brownian}
  \end{subfigure}
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{frac_pore.pdf}  % This figure might fit in earlier
  \caption{}\label{fig:frac_pore}
  \end{subfigure}
  \caption{(a) We fit the MFPT versus pore length to a function of the form $cL^{\alpha}$. 
  The MFPT grows super-linearly with increasing pore length. Solutes with lower MSDs have
  higher MFPTs. (b) If we remove correlation between hops, the MFPTs reduce significantly
  and the growth exponent stays constant at 2 for all solutes independent of dwell
  distribution parameters. Higher degrees of anti-correlation (lower $H$) lead to higher
  growth exponents. (c) Solutes that spend a large fraction of their time in the tails tend
  to have larger Hurst parameters.}\label{fig:mfpt_curves}
  \end{figure}

  Comparison of Figures~\ref{fig:mfpt_curves_regular} and~\ref{fig:mfpt_curves_brownian}
  highlights that anti-correlation between hops plays a major role in solute passage times.
  With a pore length of 50 nm, the MFPT of the fasted moving solute, ethylene glycol is 
  c.a. 5 times higher with versus without anti-correlated hops. The MFPTs of the slowest
  moving solute, acetic acid, are about 16 times higher by the same comparison. This 
  difference will only grow with increasing pore length.

  In Figure~\ref{fig:frac_pore}, we show that solutes which spend more time in the tails
  tend to have lower Hurst parameters. The densely packed, viscoelastic alkane chains 
  bounce solutes between them, severely limiting their mobility. Designing LLC membranes
  that can control or prevent the partition of solutes into the tails may offer the highest
  returns on membrane performance.

>>>>

  We determined the rate at which solutes cross macroscopic-length pores based on the
  Hill relation:~\cite{hill_free_1989}
  \begin{equation}
  J = \frac{1}{MFPT}
  \label{eqn:hill_relation}
  \end{equation}
  where $J$ is the single particle solute flux and MFPT refers to the mean first passage
  time. For particle concentrations higher than 1 at the pore entrance, one can multiply
  Equation~\ref{eqn:hill_relation} by the total number of particles to get the total flux.
  In the context of our work, the MFPT describes the average length of time it takes a particle
  to move from the pore entrance to the pore exit.

@@

  To derive an analytical equation, first consider an initial pulse spreading out 
  over time with a fixed mean. We can solve for the time-dependent probability density
  of particle positions by solving the one dimensional diffusion equation:
  \begin{equation}
  \frac{\partial p}{\partial t} = D \frac{\partial^2 p}{\partial z^2}
  \end{equation}
  The appropriate initial and boundary conditions are:
  $$BC1: t > 0, z = \infty, p = 0$$
  $$BC2: t > 0, z = 0, \frac{\partial p}{\partial z} = 0$$
  $$IC: t = 0, c = M\delta(z)$$
  where $M$ is the total number of particles in the system. It has been shown 
  elsewhere that the solution to this equation is:~\cite{cussler_diffusion:_2009}
  \begin{equation}
  p(z, t) = \frac{1}{\sqrt{4 \pi D t}}exp\bigg(\frac{-z^2}{4Dt}\bigg)
  \end{equation}
  We can make the substitution $z = z - vt$, where $v$ represents a constant average
  velocity, in order to linearly shift the mean as a function of time:
  \begin{equation}
  p(z, t) = \frac{1}{\sqrt{4 \pi D t}}exp\bigg(\frac{-(z - vt)^2}{4Dt}\bigg)
  \end{equation}
  One can track the fraction of particles, $F$, that have crossed the pore 
  boundary by integrating:
  \begin{equation}
  F(t) = \int_L^\infty p~dz = erfc\bigg(\frac{L - vt}{2\sqrt{D t}}\bigg)
  \end{equation}
  where $L$ is the pore length. This represents the cumulative first passage 
  time distribution so we take its derivative in order to arrive at the first
  passage time distribution:
  \begin{equation}
  P(t) = -\frac{1}{\sqrt{\pi}}e^{-(L - vt)^2 / (4Dt)}\bigg(-\frac{D(L - vt)}{4(Dt)^{3/2}} - \frac{v}{2\sqrt{Dt}}\bigg)
  \label{eqn:passage_times}
  \end{equation} 
  where the only free parameters for fitting are $v$ and $D$. We calculated the
  expected value of Equation~\ref{eqn:passage_times} in order to get the MFPT.

  We used the ratio of solute fluxes in order to determine membrane selectivity, $S$,
  towards solutes. Selectivity of solute $i$ versus $j$ is defined in terms of the
  ratio of solute permeabilities, $P$:~\cite{guo_pervaporation_2004}
  \begin{equation}
  S_{ij} = \frac{P_i}{P_j}
  \end{equation}
  We can relate this to solute flux using Kedem and Katchalsky's equations for 
  solvent volumetric flux, $J_v$, and solute flux, $J_s$:~\cite{kedem_permeability_1963}
  %al-zoubi_rejection_2007 clarifies that these equations can be used for nanofiltration, ultrafiltration etc. by treating the membrane as a black box subject to irreversible thermodynamics.
  \begin{equation}
  J_v = L_p(\Delta P - \Delta \pi)
  \end{equation} 
  \begin{equation}
  J_s = P_s \Delta C + (1 - \sigma)CJ_v
  \label{eqn:solute_flux}
  \end{equation}
  where $L_p$ is the pure water permeability, $\Delta P$ and $\Delta \pi$ are the 
  trans-membrane hydraulic and osmotic pressure differences, $\sigma$ is the reflection
  coefficient, $P_s$ is the solute permeability, $\Delta C$ is the trans-membrane
  solute concentration difference and $C$ is the mean solute concentration. Since our
  simulations do not include convective solute flux, we eliminate the second term 
  of Equation~\ref{eqn:solute_flux} which allows us to derive a simple expression
  for selectivity in terms of solute flux:
  \begin{equation}
  S_{ij} = \frac{J_i / \Delta C_i}{J_j / \Delta C_j}
  \label{eqn:selectivity}
  \end{equation}

%    $$ A_i = \frac{J}{(\Delta P - \Delta \pi)} $$
%  $$ S = \frac{A_i}{A_j} for a given pore length, L$$
%  $$J = A(\Delta P - \Delta \pi)$$
%  $$A = \frac{P_i}{L} $$
%  $$P_i = \frac{flux}{drivingforce \cdot L} [=] \frac{particles}{s \cdot bar \cdot L}$$

@@

  %BJC4: another (bad) way to visualize data
%  \begin{figure}
%  \centering
%  \begin{subfigure}{0.45\textwidth}
%  \includegraphics[width=\linewidth]{1mode_radar.pdf}
%  \caption{}\label{fig:1mode_radar}
%  \end{subfigure}
%  \begin{subfigure}{0.45\textwidth}
%  \includegraphics[width=\linewidth]{2mode_radar.pdf}
%  \caption{}\label{fig:2mode_radar}
%  \end{subfigure}
%  \caption{Another way to visualize Tables~\ref{table:sfbm_params} and~\ref{table:sfbm_params_2mode}}\label{fig:radars}
%  \end{figure}
%  
%  \begin{figure}
%  \centering
%  \begin{subfigure}{0.45\textwidth}
%  \includegraphics[width=\linewidth]{1mode_radar_nolines.pdf}
%  \caption{}\label{fig:1mode_radar}
%  \end{subfigure}
%  \begin{subfigure}{0.45\textwidth}
%  \includegraphics[width=\linewidth]{2mode_radar_nolines.png}   % for some reason it has trouble saving this as a PDF
%  \caption{}\label{fig:2mode_radar}
%  \end{subfigure}
%  \caption{Same as Figure~\ref{fig:radars} but with no lines between points}\label{fig:radars_nolines}
%  \end{figure}

@@

  \subsection{Solute Flux}\label{section:mfpt}

  We used the 1-mode sFBMcut (see Table~\ref{table:anomalous_models}) anomalous 
  diffusion model in order to demonstrate how one can use its realizations in
  order to calculate the flux (section~\ref{method:mfpt}) of solutes given model 
  parameters extracted from MD simulations. We believe the 1-mode sFBMcut model
  shows the most promising predictive ability of the models studied in this work. 
  %Any approach which allows one to 
  %generate realizations of solute trajectories can be used in place of the AD model.

  %BJC4: not sure if the next couple paragraphs belong in methods
  It is computationally unfeasible to simulate trajectories long enough that they
  traverse the length of a macroscopic pore. To date, the thinnest H\textsubscript{II}
  LLC membrane synthesized with the monomer in this work was 7$\mu$m thick. Using
  24 cores to simulate trajectory realizations in parallel, it takes on the order 
  of 1 day to simulate 50000 realizations of solutes traversing a 50 nm pore. The 
  RAM requirements and performance scales greater than linearly and thus would 
  take an unfeasible amount of memory and time to simulate transport through a 
  pore over 100 times longer.

  %BJC4: methods?
  We used simulated trajectories which traverse computationally-reasonable length
  pores in order to construct an empirical model which one can use to estimate 
  particle flux for arbitrary length pores. As shown in Figure~\ref{fig:flux_curves},
  the flux appears to scale according to a power law of the form:
  \begin{equation}
  J(L) = cL^{-\beta} 
  \label{eqn:flux_decay}
  \end{equation}
  %while the MFPT scales as its inverse, $(1/c)L^{\beta}$.

  The prefactor, $c$, in Equation~\ref{eqn:flux_decay} grows with increasing
  solute MSD while the value of the growth exponent, $\beta$, is inversely 
  related to the Hurst parameter, $H$ (see Figure~\ref{fig:flux_parameters}).
  In Figure~\ref{fig:flux_curves_brownian}, we removed the correlation 
  structure (set $H$=0.5) from AD trajectory realizations so that all hops 
  were uncorrelated. Regardless of the parameters of the dwell time distribution,
  the growth exponent is always about 2, the same value that we observe for 
  Brownian motion (see Figure~\ref{S-fig:mfpt_curve_brownian} of the 
  Supporting Information).

  Comparison of Figures~\ref{fig:flux_curves_regular} and~\ref{fig:flux_curves_brownian}
  highlights that anti-correlation between hops plays a major role in solute flux.
  With a pore length of 50 nm, the flux of the fasted moving solute, ethylene glycol is 
  c.a. 5 times lower with versus without anti-correlated hops. The flux of the slowest
  moving solute, acetic acid, is about 16 times lower by the same comparison. This 
  difference will only grow with increasing pore length. Note also that the ranking
  of solute fluxes changes. Methanol jumps from the second lowest to the highest
  flux when we remove hop correlation. The high degree of anti-correlation among
  hops made by methanol outweighs its relatively wide hop length distribution.

%   In Figure~\ref{fig:mfpt_curves_regular} we show that the MFPT grows super-linearly
%  with pore length. We fit a curve of the form  to the curves in order
%  to estimate the growth exponent, $\alpha$. The growth exponent is inversely proportional
%  to the solute MSDs which is consistent with longer passage times for solutes that
%  progress through the pores more slowly.

  %BJC: need simulations with increased length for L=35-50 nm pores of acetic acid and methanol
  %BJC: see of clauset has better way of fitting t^-alpha (alpha might be too big)
  %BJC: not sure mfpt is necessary. Takes up space.
  %BJC: change alpha to beta
  %BJC: make a chart with parameters
  %BJC: should have a paragraph about c -- roughly follows their MSDs.
  \begin{figure}
  \centering
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{flux_curves.pdf}
  \caption{Fits to $cL^{-\beta}$}\label{fig:flux_curves_regular}
  \end{subfigure}
  % BJC: could use better color scheme. Maybe outline hurst parameters
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{flux_parameters.pdf}
  \caption{}\label{fig:flux_parameters}
  \end{subfigure}
%  \begin{subfigure}{0.475\textwidth}
%  \includegraphics[width=\textwidth]{mfpt_curves.pdf}
%  \caption{}\label{fig:mfpt_curves}
%  \end{subfigure}
  %BJC: going out to 3 decimal places seems silly
  %BJC: might get closer to beta=2 if I use different curve fitting technique than fitting line to log-log
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{flux_curves_brownian.pdf}
  \caption{Fixed $H=0.5$}\label{fig:flux_curves_brownian}
  \end{subfigure}
%  \begin{subfigure}{0.475\textwidth}
%  \includegraphics[width=\textwidth]{mfpt_curves_brownian.pdf}
%  \caption{}\label{fig:mfpt_curves_brownian}
%  \end{subfigure}
  \caption{(a) The single particle solute flux decays with increasing pore length. 
  Solutes with high MSDs have higher fluxes. (b) We fit the single particle solute flux 
  versus pore length, $L$, to a power law function of the form $cL^{-\beta}$ (dashed 
  lines in (a)). Solutes with higher fluxes have higher values of $c$. Solutes with 
  smaller Hurst parameters, $H$, have larger values of $\beta$. (c) If we remove 
  correlation between hops (set $H$=0.5), the flux increases significantly and 
  $\beta$ is constant at about 2 for all solutes independent of dwell distribution
  parameters.}\label{fig:flux_curves}
  \end{figure}

  % Partially in methods?
  The power law decay of the flux with pore length implies the following relationship
  for selectivity via substitution of Equation~\ref{eqn:flux_decay} into 
  Equation~\ref{eqn:selectivity}:
  \begin{equation}
  S_{ij}(L) = \bigg(\frac{c_i}{c_j}\bigg)L^{(\beta_i - \beta_j)}
  \label{eqn:selectivity_ratio}
  \end{equation}

  In Figure~\ref{fig:selectivity}, we plot Equation~\ref{eqn:selectivity_ratio} for pore
  lengths ranging from those studied in Figure~\ref{fig:flux_curves} to macroscopic-length
  pores. Greater differences in $\beta$ values leads to selectivities that are stronger 
  functions of pore length. For $(c_i / c_j)=1$, LLC membranes will be more selective
  towards passage of solutes with lower $\beta$ values versus those with high $\beta$ values.
  When $\beta$ values are equal, as is the case for uncorrelated motion, selectivity only
  depends on the pre-factors, $c$.

  \begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{selectivity.pdf}
  \caption{The selectivity of 2 species changes monotonically with pore length. The
  strength of dependence on pore length depends on the difference between $\beta$ values.}
  \label{fig:selectivity}
  \end{figure}

  Out of the solutes studied, the data suggests that this particular LLC membrane 
  would be most useful for selectively separating ethylene glycol from methanol and 
  acetic acid. 
  %It also shows reasonable selectivity of Urea over acetic acid.  
  Ethylene glycol has the highest $c$ value and the lowest $\beta$ value. The high
  value of $c$ itself greatly enhances its selective passage. Even more importantly,
  the relatively large difference between its $\beta$ value and those of methanol 
  and acetic acid give rise to a strong pore length dependence. The selectivity is
  enhanced as the pore length increases at the cost of lower solute flux, a 
  phenomenon ubiquitous in polymer separations membranes.~\cite{geise_water_2011}

  Of course, this approach to modeling LLC membrane selectivity neglects a number of
  factors that influence real-world aqueous membrane separations. The most glaring is 
  that we neglected convective flux which should work to increase overall solute flux 
  and lower selectivity (see Equation ~\ref{eqn:solute_flux}). Defects in the pore 
  structure will have a strong influence on solute flux which may exacerbate the length 
  dependence of flux. Even in a perfectly structured membrane, the barrier to pore
  entry from the bulk solution will have its own influence on experimentally observed 
  selectivities. This energetic barrier can be made even worse by foulants deposited on
  the membrane's surface. Therefore, our predictions should be primarily used to 
  identify the most promising types of selective separations for an H\textsubscript{II}
  LLC membrane made from a specific monomer.
  %BJC: though about mentioning fouli

    Neglects concentration polarization, membrane defects and fouling.

%  BJC4: Uncomment to see MFPT curves
%  \begin{figure}
%  \centering
%  \begin{subfigure}{0.325\textwidth}
%  \includegraphics[width=\textwidth]{mfpt_curves.pdf}
%  \caption{Correlated Hops}\label{fig:mfpt_curves_regular}
%  \end{subfigure}
%  \begin{subfigure}{0.325\textwidth}
%  \includegraphics[width=\textwidth]{mfpt_curves_brownian.pdf}
%  \caption{Uncorrelated Hops}\label{fig:mfpt_curves_brownian}
%  \end{subfigure}
%  \begin{subfigure}{0.325\textwidth}
%  \includegraphics[width=\textwidth]{frac_pore.pdf}  % This figure might fit in earlier
%  \caption{}\label{fig:frac_pore}
%  \end{subfigure}
%  \caption{(a) We fit the MFPT versus pore length to a function of the form $cL^{\alpha}$. 
%  The MFPT grows super-linearly with increasing pore length. Solutes with lower MSDs have
%  higher MFPTs. (b) If we remove correlation between hops, the MFPTs reduce significantly
%  and the growth exponent stays constant at 2 for all solutes independent of dwell
%  distribution parameters. Higher degrees of anti-correlation (lower $H$) lead to higher
%  growth exponents. (c) Solutes that spend a large fraction of their time in the tails tend
%  to have larger Hurst parameters.}\label{fig:mfpt_curves}
%  \end{figure}

%  In Figure~\ref{fig:frac_pore}, we show that solutes which spend more time in the tails
%  tend to have lower Hurst parameters. The densely packed, viscoelastic alkane chains 
%  bounce solutes between them, severely limiting their mobility. Designing LLC membranes
%  that can control or prevent the partition of solutes into the tails may offer the highest
%  returns on membrane performance.
